---
title: "final_project"
author: "Nathan Hepler"
date: "12/01/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)
library(agricolae)
library(stringr)

## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project')
```

### The Data

The American Time Use Survey (ATUS) is a time-use survey of Americans, which is sponsored by the Bureau of Labor Statistics (BLS) and conducted by the U.S. Census Bureau. Respondents of the survey are asked to keep a diary for one day carefully recording the amount of time they spend on various activities including working, leisure, childcare, and household activities. The survey has been conducted every year since 2003.

Included in the data are main demographic variables such as respondents' age, sex, race, marital status, and education. The data also includes detailed income and employment information for each respondent. While there are some slight changes to the survey each year, the main questions asked stay the same. You can find the data dictionaries for each year on [https://www.bls.gov/tus/dictionaries.htm](https://www.bls.gov/tus/dictionaries.htm)


### Accessing the Data

There are multiple ways to access the ATUS data; however, for this project, you'll get the raw data directly from the source. The data for each year can be found at [https://www.bls.gov/tus/#data](https://www.bls.gov/tus/#data). Once there, there is an option of downloading a multi-year file, which includes data for all of the years the survey has been conducted, but **for the purposes of this project, let's just look at the data for 2016**. Under **Data Files**, click on `American Time Use Survey--2016 Microdata files`. 

You will be brought to a new screen. Scroll down to the section **2016 Basic ATUS Data Files**. Under this section, you'll want to **click to download** the following two files: `ATUS 2016 Activity summary file (zip)` and `ATUS-CPS 2016 file (zip)`. 

* `ATUS 2016 Activity summary file (zip)` contains information about the total time each ATUS respondent spent doing each activity listed in the survey. The activity data includes information such as activity codes, activity start and stop times, and locations.
* `ATUS-CPS 2016 file (zip)` contains information about each household member of all individuals selected to participate in the ATUS.

Once they've been downloaded, you'll need to **unzip the files**. Once unzipped, you will see the dataset in a number of different file formats including `.sas`, `.sps`, and `.dat` files. **We'll be working with the .dat files.**

### Loading the Data into R

Use the first approach explained above to download and access the ATUS data for 2016. Download the CPS and Activity Summary files in a folder and unzip them and within each folder upload the files ending in .dat to `data/raw_data` filder on RStudio.cloud. To load the data in, **run the code in the `atus-data` code chunk** to create an object called `atus.all`.

### Importing data

```{r atus-data, warning=FALSE}
atus.cps <- read.delim('final_project/data/raw_data/atuscps_2016.dat', sep=",")
atus.sum <- read.delim('final_project/data/raw_data/atussum_2016.dat', sep=",")
atus.all <- atus.sum %>%
  left_join(atus.cps %>% filter(TULINENO==1), by = c("TUCASEID"))
```

### Exploratory Analysis of Child Care Data

```{r initial-exploration}
### Add Code Here
socializing <- atus.all %>% 
  select(t120101)

mean(socializing$t120101)
```


```{r creating-childcare-var}
## columns with activity code starting with t0301 are all associated with “Caring For & Helping HH Children”.
## we want to sum those up to generate a column containing time spent collectively on 'CHILDCARE'
## use rowSums function alongside grep to:
##    1. Find all columns containing 't0301' (grep)
##    2. Sum up the total rows so that the # of obs. is consistent with 
##       dataset as a whole
atus.all <- atus.all %>% 
    mutate(CHILDCARE = rowSums(.[, grep('t0301', colnames(.))]))
```

```{r childcare-density-plot}
ggplot(atus.all, aes(CHILDCARE)) +
    geom_density() + 
  labs(title='Average time spent on childcare',
       subtitle='American Time Use Survey (2016)',
       caption='Source: Bureau of Labor Statistics (BLS)') + 
  xlab(label='Minutes')
```

```{r gender-analysis}
atus.all %>% 
    group_by(TESEX) %>% 
    summarise(n=mean(CHILDCARE))
```

```{r replacing-na}
## replace -1 in the variable TRDPFTPT with NA.

## first way i thought of: 
## new <- function(column){
##  for(val in 1:length(column)){
##    if(column[val] == -1){
##      column[val] = 'NA'
##    }
##  }
##  return(column)
##}
##NAs <- new(atus.all$TRDPFTPT)
##atus.all$TRDPFTPT <- NAs

## easier way:
atus.all$TRDPFTPT <- sub(-1, 'NA', atus.all$TRDPFTPT)
```

```{r exploratory-analysis}
## add your exploratory analysis code here
## all filtered to make sure HAVE CHILDREN (TRCHILDNUM > 0)

## age v childcare time
atus.all %>% 
  filter(TRCHILDNUM > 0) %>% 
  select(TEAGE, CHILDCARE) %>% 
  arrange(TEAGE) %>% 
  filter(TEAGE > -1) %>% 
  ggplot(aes(x=as.numeric(TEAGE), y=CHILDCARE)) + 
  geom_point() + 
  ylab(label='Time spent on childcare (minutes)') + 
  xlab(label='Age of parent (years)') + 
  labs(title='Relationship between parent age and childcare',
       subtitle='\tSee more of a bell curve, with a general peak in mid-30s')

## pay v childcare time

## running stats
pay.aov <- aov(CHILDCARE~as.character(HEFAMINC), data=atus.all %>%
                 filter(TRCHILDNUM > 0))
pay.HSD <- HSD.test(pay.aov, trt='as.character(HEFAMINC)', alpha=0.05)

## plotting
atus.all %>% 
  filter(TRCHILDNUM > 0) %>% 
  select(HEFAMINC, CHILDCARE) %>% 
  arrange(HEFAMINC) %>%
  ggplot(aes(x=HEFAMINC, y=CHILDCARE)) + 
  geom_boxplot(aes(group=HEFAMINC)) + 
  ylab(label='Time spent on childcare (minutes)') + 
  xlab(label='Household salary') + 
  labs(title='Relationship between income and childcare',
       subtitle='\tNo clear relationship observed')

## marital status v childcare
## converting numbers to text
atus.all$PEMARITL <- sub(1, 'Married - spouse present', atus.all$PEMARITL)
atus.all$PEMARITL <- sub(2, 'Married - spouse absent', atus.all$PEMARITL)
atus.all$PEMARITL <- sub(3, 'Widowed', atus.all$PEMARITL)
atus.all$PEMARITL <- sub(4, 'Divorced', atus.all$PEMARITL)
atus.all$PEMARITL <- sub(5, 'Separated', atus.all$PEMARITL)
atus.all$PEMARITL <- sub(6, 'Never married', atus.all$PEMARITL)

## running stats
mar.aov <- aov(CHILDCARE~PEMARITL, data=atus.all %>%
                 filter(TRCHILDNUM > 0))
mar.HSD <- HSD.test(mar.aov, trt='PEMARITL', alpha=0.05)

## plotting as table, including a column for stats
atus.all %>% 
  filter(TRCHILDNUM > 0) %>% 
  select(Status=PEMARITL, CHILDCARE) %>% 
  group_by(Status) %>% 
  summarise(Average=mean(CHILDCARE), SD=sd(CHILDCARE)) %>% 
  arrange(desc(Average)) %>% 
  mutate(ANOVA=mar.HSD$groups[,2]) %>% 
  kable(digits=2, format='pandoc', align='rccc', caption='Average time spent on childcare by marital status')


## fulltime v partime and childcare
## converting numbers to labels
atus.all$TRDPFTPT <- sub(1, 'Full time', atus.all$TRDPFTPT)
atus.all$TRDPFTPT <- sub(2, 'Part time', atus.all$TRDPFTPT)

## doing stats
job.test <- t.test(CHILDCARE~TRDPFTPT, data=atus.all %>% 
                 filter(TRCHILDNUM > 0, TRDPFTPT == 'Full time' | 
                          TRDPFTPT == 'Part time'))

## plotting
atus.all %>% 
  filter(TRCHILDNUM > 0, TRDPFTPT == 'Full time' | TRDPFTPT == 'Part time') %>% 
  select(Status=TRDPFTPT, CHILDCARE) %>% 
  group_by(Status) %>% 
  summarise(Average=mean(CHILDCARE), SD=sd(CHILDCARE)) %>% 
  kable(digits=2, format='pandoc', align = 'rcc', caption=paste0('Average time spent on childcare by employmeent status.\np-value: ', signif(job.test$p.value, digits=3)))


## sex v childcare
## doing stats
sex.test <- t.test(CHILDCARE~as.character(TESEX), data=atus.all %>% 
                 filter(TRCHILDNUM > 0))
## plotting
atus.all %>% 
  filter(TRCHILDNUM > 0) %>% 
  select(TESEX, CHILDCARE) %>% 
  group_by(TESEX) %>% 
  ggplot(aes(x=as.character(TESEX), y=CHILDCARE)) + 
  geom_boxplot(aes(group=TESEX)) + 
  scale_x_discrete(labels=c('Male', 'Female')) + 
  labs(title='Average time spent on childcare by gender',
       subtitle='\tWomen spend more time with children',
       caption = paste0('p-value for Welch t-test: ', signif(sex.test$p.value, digits=3))) + 
  xlab(label='Gender') + 
  ylab(label='Time spent on childcare (min)')

  
  
```

### Regression Analysis

```{r reg-analysis}
## add your regression analysis code here
reg_model <- lm(CHILDCARE~PEMARITL + TEAGE + TESEX + TRCHILDNUM + TRDPFTPT,
                data=atus.all %>% 
                  filter(TRCHILDNUM >= 1))
summary(reg_model)
```

### Exploratory Analysis of Age and Activities

```{r activity-cats}
atus.wide <- atus.all %>%
    mutate(act01 = rowSums(.[, grep('t01', colnames(.))]),
           act02 = rowSums(.[, grep('t02', colnames(.))]),
           act03 = rowSums(.[, grep('t03', colnames(.))]),
           act04 = rowSums(.[, grep('t04', colnames(.))]),
           act05 = rowSums(.[, grep('t05', colnames(.))]),
           act06 = rowSums(.[, grep('t06', colnames(.))]),
           act07 = rowSums(.[, grep('t07', colnames(.))]),
           act08 = rowSums(.[, grep('t08', colnames(.))]),
           act09 = rowSums(.[, grep('t09', colnames(.))]),
           act10 = rowSums(.[, grep('t10', colnames(.))]),
           act11 = rowSums(.[, grep('t11', colnames(.))]),
           act12 = rowSums(.[, grep('t12', colnames(.))]),
           act13 = rowSums(.[, grep('t13', colnames(.))]),
           act14 = rowSums(.[, grep('t14', colnames(.))]),
           act15 = rowSums(.[, grep('t15', colnames(.))]),
           act16 = rowSums(.[, grep('t16', colnames(.))]),
           # act17 = , there is no category 17 in the data
           act18 = rowSums(.[, grep('t18', colnames(.))])) %>% 
    select(TUCASEID, TEAGE, HEFAMINC, starts_with("act"))
```

```{r wide-to-long}
## Activity will be name of column with acts and mins column with values (observations)
atus.long <- atus.wide %>% 
  reshape2::melt(variable.name='ACTIVITY', value.name='MINS', id.vars=c('TUCASEID', 'TEAGE', 'HEFAMINC'))

```

```{r age-activity}
atus.long %>% 
    group_by(ACTIVITY, TEAGE) %>% 
    summarise(AVGMINS = mean(MINS)) %>% 
  ggplot(aes(x=TEAGE, y=AVGMINS)) + 
  geom_bar(stat='identity') + 
  facet_wrap(~ACTIVITY)
```

### Exploratory Analysis of Income and Activities

```{r activity-income}

## expand in scale_y removed spacing between axis and bars
## guide_colourbar allows for adjustment of bar size

p1<-atus.long %>% 
  filter(HEFAMINC > 0) %>% 
  group_by(ACTIVITY, HEFAMINC) %>% 
  arrange(HEFAMINC) %>% 
  ggplot(aes(x=ACTIVITY, y=MINS, fill=HEFAMINC)) + 
  geom_bar(position='fill', stat='identity') + 
  coord_flip() + theme_classic() + 
  scale_y_discrete(expand=c(0,0), breaks = 1:17, labels= act_labels) + 
    labs(title="Income distributions in Americans' pastimes",
       subtitle='This chart shows the income distribution of adult participants in different\npastimes.') + 
  scale_fill_gradientn(colours=c('darkblue','steelblue3', 'deepskyblue3', 'darkslategray1','lightblue','goldenrod1', 'gold3', 'gold4', 'darkgoldenrod4'), 
                       limits=c(1,16), 
                       breaks=seq(1,16, by=3), 
                       labels=c('<5k', '10k', '20k', '35k', '60k', '150k+'),
                       guide='colourbar',
                       na.value = 'green',
                       name='Total population\n\nYearly household income') + 
  guides(fill=guide_colorbar(barwidth = 19,
                             barheight = 1)) + 
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = 'bottom', 
    legend.title = element_text(size=7, vjust=0.8),
    legend.text = element_text(size=7),
    legend.title.align = 1,
    plot.title = element_text(face='bold')
  )
p1
```


```{r save-plot}
## save the plot above
ggsave(plot=p1, filename='activity-income.png',
       path='final_project/figures/explanatory_figures')
```